Pipit: Reactive Systems in \fstar ~(Extended Abstract)

### 2023-05-11 2023pipit/Main.tex
almost

In this extended abstract, we will briefly describe the following preliminary results, which we intend to publish pending further evaluation:
\begin{itemize}
  \item We identify the invariants and lemmas required to prove correctness of the translation from a clock-free Lustre-style language to a transition system (\autoref{ss:meta-theory}).
  \item We present a syntax-directed operational semantics for a clock-free Lustre-style language (\autoref{ss:operational-semantics}).
    % TODO: CUT WAFFLE
    This syntax-directed semantics is in contrast to previous \emph{relational} operational semantics of Lustre, which are not syntax-directed and require detailed proofs of determinism.
    Proving determinism on our semantics requires only straightforward induction; we believe that this semantics may be better for proving other properties too.
  \item We show preliminary results that demonstrate the effectiveness of Pipit by implementing a verified controller for a simple embedded system (\autoref{s:coffee-machine}).
\end{itemize}

@@

\section{Pipit in flight}
\label{s:tut}

@@

Although this definition is simple, verifying it \emph{and} executing it in Lustre is somewhat impractical, as tool support varies widely.
To write this program in Lustre, we would likely need to reformulate the program to use arrays.
Vélus does not support arrays; Kind2 uses a custom syntax for arrays that no compiler supports; and Lustre v6 does support compiling arrays and recursive nodes, but its model-checker, Lesar~\cite{raymond2008synchronous} does not.

Instead, we take the embedded language approach of defining a simple core language inside a theorem prover (\fstar{}).
By implementing Pipit as an embedded language, we can express the above FIR filter, but the lists are evaluated away by the meta-language to a simple core program which can be easily executed in statically-bounded runtime and space.
For example, we can compute the weighted moving average with two coefficients:

@@

To state properties about our filter, it is useful to be able to state temporal properties.
We do not need a separate specification language; instead, we can implement computable variants of temporal operators from \emph{past-time} linear temporal logic \cite{halbwachs1993executable,lichtenstein1985glory}.
We name the past-globally operator \emph{sofar}, as in \emph{the property has been true so far}:

@@

We can now state a \emph{bounded-input-bounded-output} (BIBO) property, which says that if the inputs have always been within some particular range, then the outputs are also within the range.
Such properties can occur as an internal invariant in a larger proof.

@@

This property states that, for this particular set of coefficients, if the input has always been in the range $[-n, n]$, then the output is also within the range $[-n, n]$.
Note that the upper bound $n$ is a nonnegative real rather than a stream of reals, which means that $n$ stays constant across the whole stream.
To prove that this property holds, we translate to a transition system.
In this case, simple induction over the transition relation is sufficient to prove the property.
There is some boilerplate required to perform the induction, but the step case and base case are automatically proved by \fstar{}:

@@

If we wish to prove a similar BIBO property for three or more coefficients, this standard induction over the transition system is no longer sufficient, and we need a stronger invariant.
One simple automated way to strengthen the invariants is via \emph{k-induction}~\cite{hagen2008scaling}, which looks at multiple steps of the transition relation at a time.
For example, we can define analogous functions $\fir_3$ and $\bibo_3$ which operate on the coefficients $[0.7; 0.2; 0.1]$, and use k-induction for $k = 2$:

@@

% The issue is that each delayed values inside the FIR filter becomes a separate state variable in the transition system.
% When looking at the result of a single step of the transition relation in isolation, we can prove that $(\xfby{0}{\ii}) \le n$ is equivalent to $\xfby{\top}{(\ii \le n)}$.
% When nesting delays, however, the same sort of distributive property is not clear from just a single step of the transition relation; each delay gets its own state variable, and the relationship between the different state variables is lost.
% We really need a stronger invariant to do this.

% \TODO{Required for submission: writing}
% \begin{itemize}
%   \item introduction
%   \item running example: FIR?
%   \item describe translation to transition system? detail probably won't fit, translation invariant is the most interesting part.
% \end{itemize}
% \TODO{Required for submission: implementation}
% \begin{itemize}
%   \item finish proof of system correctness
%   \item state end-to-end / capstone theorem that k-inductively verified systems apply to original semantics
%   \item implement unverified CSE pass
%   \item verify running example, with CSE and k-induction
% \end{itemize}

% Cite: Kind2 \cite{champion2016kind2}, Lustre \cite{caspi1995functional}, Vélus \cite{bourke2017formally}.
% Cite: the original Kind \cite{hagen2008scaling}.
% \CITEME{2022 thesis on Agda semantics of Lustre?}
% Past-time modal temporal logic \cite{lichtenstein1985glory}
% Lustre support for temporal logic \cite{halbwachs1993executable}

\section{Core language}
\label{s:core}

\input{figures/core-grammar.tex}
% \input{figures/core-typing.tex}
\input{figures/core-bigstep.tex}

The grammar of Pipit is defined in \autoref{f:core-grammar}.
The expression form $e$ includes standard syntax for values ($v$), variables ($x$) and applications ($e~e'$); however, it does not include any form for defining functions other than reusing closed functions from the \fstar{} meta-language.
Most of the expression forms were introduced informally in \autoref{s:tut} and correspond to the clock-free primitives in Lustre \cite{caspi1995functional}.
The expression syntax for delayed streams ($\xfby{v}{e}$) denotes the previous value of the stream $e$, with an initial value of $v$ when there is no previous value.
Streams can be composed together using the \emph{then} notation ($\xthen{e}{e'}$) which denotes that the value of stream $e$ is used for the very first step, followed by the latest values from stream $e'$ for subsequent steps.

\subsection{Operational semantics}
\label{ss:operational-semantics}

The operational semantics, expressed as a bigstep relation, are defined in \autoref{f:core-bigstep}.
The judgment form $\bigstep{\Sigma}{e}{v}$ takes a \emph{stream history} ($\Sigma$), which is a non-empty sequence of stores, as well as an expression to evaluate ($e$) and the final value ($v$).

Rule Value denotes that a value evaluates to itself under any streaming history.
Rule Var denotes that evaluating a variable requires looking it up in the right-most (most recent) store; the syntax $\Sigma_\bot; \sigma$ means that this rule applies to any streaming history with one or more steps.
Rule App-Meta allows applying functions on the meta-level; the syntax $(\mlamX{}) v'$ refers to meta-level application of some meta-term.

The rules $\mbox{Fby}_1$ and $\mbox{Fby}_S$ describe the semantics of delayed streams.
Rule $\mbox{Fby}_1$ denotes that when a delay $\xfby{v}{e}$ is evaluated in a context with exactly one step --- that is, at the very start of program execution --- it simply returns the initial value $v$.
Rule $\mbox{Fby}_S$ applies when the same delay is evaluated in a context with strictly more than one step ($\Sigma; \sigma$), in which case, the most recent step is discarded, and the original expression $e$ is evaluated under the previous history.

Next, rules ($\xthenarrow_1$) and ($\xthenarrow_S$) describe the semantics of streams composed by \emph{then}-expressions; as with the rules for delayed streams, the first rule applies when the streaming history contains exactly one step, while the second applies for histories of strictly more than one step.

Rule ($\mu{}$) describes the semantics of recursive streams.
This rule performs a single unfolding of a recursive expression $\xrec{x}{e[x]}$ so that all occurrences of $x$ inside $e$ are replaced with the recursive expression itself.
This rule is not trivially terminating, but so long as any occurrences of the recursive binder are guarded by a delay, then all occurrences of the recursive binder will have a shorter stream history.
This intuition is formalised in \REF{}.

Rule Let uses a standard substitution-based semantics for let-expressions.
Finally, rule Check evaluates a property and requires it to be true ($\top$).

The semantics here, particularly the recursive rule, differ from previous semantics of Lustre.
In the reactive semantics of \citet{caspi1995functional}, as well as the coinductive semantics of Vélus \cite{bourke2017formally} (\CITEME{find best paper}), the semantics for mutually recursive definitions is locally non-deterministic, in that the semantics does not define which particular order mutual definitions are evaluated in.
This local non-determinism complicates certain proofs, even though the overall evaluation \emph{is} deterministic \TODO{vague}.

In contrast, the semantics presented here is syntax-directed and completely deterministic.
We have a mechanised proof of variant of progress theorem for well-formed (\emph{causal}) programs as well as closure under substitution and determinism, which is described in \autoref{ss:meta-theory}.
However, the substitution-based semantics here can duplicate expressions, and encoding mutually recursive definitions requires multiple substitutions.
For this reason, the semantics here is not suitable for actual execution.

\subsection{Causality}
\label{ss:causality}

Talk about the causality restriction.
Does expression $e$ have a dependency on the most-recent value of expression $e'$?

\begin{code}
let rec direct_dependency (e e': exp) : bool =
  if e = e' then true
  else
  match e with
  | XVal _ -> false
  | XVar x' -> false
  | XPrim2 _ e1 e2 -> direct_dependency e1 e' || direct_dependency e2 e'
  | XIte ep et ef -> direct_dependency ep e' || direct_dependency et e' || direct_dependency ef e'
  | XThen e1 e2 -> direct_dependency e1 e' || direct_dependency e2 e'
  | XPre _ -> false
  | XMu e1 -> direct_dependency e1 (lift e' 0)
  | XLet e1 e2 -> direct_dependency e1 e' || direct_dependency e2 (lift e' 0)

let rec causal (e: exp): bool =
  match e with
  | XVal _ -> true
  | XVar x' -> true
  | XPrim2 _ e1 e2 -> causal e1 && causal e2
  | XIte ep e1 e2 -> causal ep && causal e1 && causal e2
  | XThen e1 e2 -> causal e1 && causal e2
  | XPre e1 -> causal e1
  | XMu e1 -> causal e1 && not (direct_dependency e1 (XVar 0))
  | XLet e1 e2 -> causal e1 && causal e2
\end{code}

\begin{lemma}[no dependency, no difference]
If $e$ does not have a direct depend on $x$, then any evaluation of $e$ cannot depend on the current value of $x$.

For all expressions $e$ and well-scoped variables $x$,
if $\neg \mbox{directly} ~e~ x$ and
$\bigstep{\Sigma; (\sigma, x \mapsto v_x)}{e}{v}$
then for any $v_x'$, evaluation is unchanged
$\bigstep{\Sigma; (\sigma, x \mapsto v_x')}{e}{v}$
\end{lemma}

We also need:
\begin{lemma}[Bigstep recursive XMu]
  If a recursive expression $\xrec{x}{e}$ evaluates to some sequence of values $v$, then evaluating the recursive expression doesn't depend on the last value in the sequence:

  if $\bigsteps{\Sigma}{\xrec{x}{e}}{v_0 v_1 \cdots v_{n-1} v_n }$
  then for any $v'$,
  $\bigsteps{\Sigma \triangleright^{\uparrow} (v_0 v_1 \cdots v_{n-1} v')}{e}{v_0 v_1 \cdots v_{n-1} v_n}$
\end{lemma}

% let bigstep_recursive_XMu (#outer #inner: nat) (e: exp)
%   (streams: C.table (outer + 1) inner)
%   (vs: C.vector value outer) (v v': value)
%   (hBSmu: bigstep streams (XMu e) (v :: vs)):
%     bigstep (C.table_map_append (C.table_of_values (v' :: vs)) streams) e (v :: vs) =

\begin{lemma}[progress]
  if expression is causal, then for any valid stream history it should evaluate to a value:

  if $\typing{\Gamma}{e}{\tau}$ and $e \mbox{causal}$ and $\Gamma \vdash* \Sigma$, then exists $v$ such that
  $\bigstep{\Sigma}{e}{v}$ and $\mtyping{}{v}{\tau}$
\end{lemma}

\begin{lemma}[monotonic progress]
  if an expression $e$ evaluates to a sequence $vs$ under a stream history $\Sigma$ and a new stream store $\sigma$ is valid, then
  there exists some $v'$ such that
  $\bigsteps{\Sigma; \sigma}{e}{vs; v}$
\end{lemma}

\subsection{Meta-theory}
\label{ss:meta-theory}

We use a typed representation, which ensures an intrinsic variant of preservation.
For programs that satisfy the causality restriction, we have proved a variant of the progress theorem.

\begin{code}
let bigstep_deterministic
  (#streams: list (C.row 'c))
  (#e: exp 'c 'a)
  (#v1 #v2: 'a)
  (hBS1: bigstep streams e v1) (hBS2: bigstep streams e v2):
    Lemma (ensures (v1 == v2)) (decreases hBS1) =
  bigstep_proof_equivalence hBS1 hBS2
\end{code}

Monotonicity

Substitution-closure
\begin{code}
let rec bigstep_substitute_as_var (e p: exp)
  (streams1: C.table outer inner1)
  (streams2: C.table outer inner2)
  (vsp vsep: C.vector value outer)
  (hBSp: bigstep (C.table_map_append streams1 streams2) p vsp)
  (hBSep: bigstep (C.table_map_append streams1 streams2) (subst e inner1 p) vsep):
    Tot (bigstep (C.table_map_append streams1 (C.table_map_append (C.table_of_values vsp) streams2)) e vsep) (decreases hBSep) =

\end{code}

\section{Proving it with transition systems}

Describe the translation from Pipit expressions to labelled transition systems.

K-induction

Statement of LTS soundness:
\begin{code}
let system_eval_complete'
    (#outer: nat) (#vars: nat)
    (e: exp { causal e /\ wf e vars })
    (streams: C.table outer vars)
    (vs: C.vector value outer)
    (hBS: bigstep streams e vs):
      Lemma (exists (s': option (state_of_exp e)).
        system_stepn'
          (system_of_exp e vars)
          (C.Table?._0 streams) vs s') =
    match outer with
    | 0 -> assert
      (system_stepn'
        (system_of_exp e vars)
        (C.Table?._0 streams) vs None)
    | _ -> system_eval_complete e streams vs hBS
\end{code}

Tactics: \cite{martinez2019meta}.

Invariant generation: ICE \cite{garg2014ice}; template-based \cite{kahsai2011instantiation}; PDR explanation \cite{een2011efficient}; IC3 first introduced \cite{bradley2011sat}

\section{Execution}

We have implemented an unverified imperative code generation transform, by making use of \fstar{}'s existing C code extraction support \cite{protzenko2017verified}.
% This transform generates \fstar{} functions in the

This allows us to generate C code which executes in bounded space on embedded devices.

Other compilers for Lustre-style languages, such as Vélus \cite{bourke2017formally} and Heptagon \cite{gerard2012modular} ensure that the generated code can be executed in bounded memory and that each step takes a bounded time.
With a carefully chosen set of primitives, the core streaming operations shown here would ensure bounded memory and time.
For practicality, however, we currently allow embedding arbitrary total functions from the \fstar{} meta-language, which are not necessarily bounded in time.
(Is this the same as Lucid Synchrone and Zélus? \CITE)

\section{Coffee machine}
\label{s:coffee-machine}

@@

hardware language in Agda \cite{harrison2021mechanized}

Refinement types for Zélus \cite{chen2022synchronous}

\TODO{discuss TCB: big, much bigger than Vélus}

\section{Conclusion and future work}

Future work: verify imperative code generation; verify CSE; case studies: antilock braking; clocks \cite{caspi1995functional}; contracts \cite{champion2016kind2}.

% \pagebreak
% \begin{acks}
%   acknowledgement
% \end{acks}

>>>>

In this extended abstract, we will briefly describe the following preliminary results:

\begin{itemize}
  \item embedded language reduces the need for fancy features (\autoref{s:tut});
  \item normalisation and proof automation allow us to prove certain properties with minimal effort (\autoref{ss:tut:bibo});
  \item normalisation and extraction allows us to generate unverified executable code (\autoref{s:extraction});
\end{itemize}

We will also touch upon the following work-in-progress, which we intend to publish upon completion:

\begin{itemize}
  \item we describe the syntax (\autoref{s:core});
  \item a unified semantics for theorem proving and computation;
  \item verified imperative code generation;
  \item the ability to mix manual proofs with automated proofs
  \item further evaluation with the verification of a realistic controller
\end{itemize}

%CUT
% \begin{itemize}
%   \item We identify the invariants and lemmas required to prove correctness of the translation from a clock-free Lustre-style language to a transition system (\autoref{ss:meta-theory}).
%   \item We present a syntax-directed operational semantics for a clock-free Lustre-style language (\autoref{ss:operational-semantics}).
%     % TODO: CUT WAFFLE
%     This syntax-directed semantics is in contrast to previous \emph{relational} operational semantics of Lustre, which are not syntax-directed and require detailed proofs of determinism.
%     Proving determinism on our semantics requires only straightforward induction; we believe that this semantics may be better for proving other properties too.
%   \item We show preliminary results that demonstrate the effectiveness of Pipit by implementing a verified controller for a simple embedded system (\autoref{s:coffee-machine}).
% \end{itemize}

@@

% \section{Pipit in flight}
\pagebreak
\section{Programming and verifying in Pipit}
\label{s:tut}

@@

Although this definition is simple, verifying it \emph{and} executing it in Lustre is somewhat impractical, as tool support varies widely.
To write this program in Lustre, we would likely need to reformulate the program to use arrays.
Vélus does not support arrays; Kind2 uses a custom syntax for arrays that no compiler supports; and Lustre v6 does support arrays, but its model-checker Lesar~\cite{raymond2008synchronous} does not.

Instead, we take the embedded language approach of defining a simple core language inside a theorem prover (\fstar{}), inspired by Bedrock~\cite{chlipala2013bedrock}.
By implementing Pipit as an embedded language, we can express the above FIR filter, but the lists are evaluated away by the meta-language to a simple core program which can be easily executed in statically-bounded runtime and space.
For example, we can compute the weighted moving average with two coefficients:

@@

To state properties about our filter, it is useful to be able to state temporal properties.
We do not need a separate specification language; instead, we can implement computable variants of temporal operators from \emph{past-time} linear temporal logic \cite{halbwachs1993executable,lichtenstein1985glory}.
We name the past-globally operator \emph{sofar}, as in \emph{the predicate has been true so far}:

@@

\subsection{Bounded input, bounded output}
\label{ss:tut:bibo}
We can now state a \emph{bounded-input-bounded-output} (BIBO) property, which says that if the inputs have always been within some particular range, then the outputs are also within the range.
Such properties can occur as an internal invariant in a larger proof.

@@

This property states that, for this particular set of coefficients, if the input has always been in the range $[-n, n]$, then the output is also within the range $[-n, n]$.
Note that the upper bound $n$ is a nonnegative real rather than a stream of reals, which means that $n$ stays constant across the whole stream.
To prove that this property holds, we translate to a transition system.
In this case, simple induction over the transition relation is sufficient to prove the property.
There is some boilerplate required to perform the induction, but both base and step cases are automatically proved by \fstar{}:

@@

If we wish to prove a similar BIBO property for more coefficients, standard induction over the transition system is no longer sufficient; we need a stronger invariant.
One simple automated way to strengthen invariants is via \emph{k-induction}~\cite{hagen2008scaling}, which adds more context by assuming that the property holds for $k$ previous steps of the transition relation.
For example, we can define analogous functions $\fir_3$ and $\bibo_3$ which operate on the coefficients $[0.7; 0.2; 0.1]$, and use k-induction for $k = 2$:

@@

Although the properties we have shown here boil down to simple properties about linear arithmetic, we believe that this example demonstrates a promising way to use \fstar{}'s existing proof automation for verifying reactive systems.

\pagebreak
\section{Core language}
\label{s:core}

\input{figures/core-grammar.tex}
% \input{figures/core-typing.tex}
% \input{figures/core-bigstep.tex}

The grammar of Pipit is defined in \autoref{f:core-grammar}.
The expression form $e$ includes standard syntax for values ($v$), variables ($x$) and applications ($e~e'$); however, it does not include any form for defining functions other than reusing closed functions from the \fstar{} meta-language.
Most of the expression forms were introduced informally in \autoref{s:tut} and correspond to the clock-free primitives in Lustre \cite{caspi1995functional}.
% The expression syntax for delayed streams ($\xfby{v}{e}$) denotes the previous value of the stream $e$, with an initial value of $v$ when there is no previous value.
Streams can be composed together using the \emph{then} notation ($\xthen{e}{e'}$) which denotes that the value of stream $e$ is used for the first step, followed by the latest values from stream $e'$ for subsequent steps.

Recursive streams, which can refer to previous values of the stream itself, are defined using the fixpoint operator ($\xrec{x}{e[x]}$); the syntax $e[x]$ means that the variable $x$ can occur in $e$.
% To ensure that streams are productive,
As in Lustre, recursive streams can only refer to their previous values and must be \emph{guarded} by a delay: the stream $(\xrec{x}{\xfby{0}{(x + 1)}})$ is well-defined, but stream $(\xrec{x}{x + 1})$ is invalid and has no computational interpretation.
This form of recursion differs slightly from standard Lustre, which uses a set of mutually-recursive bindings.
We use this form to specify an operational semantics that is syntax-directed, as opposed to the mutually-recursive form in \citet{caspi1995functional} which is not syntax directed.
The syntax-directed semantics simplifies the proof of determinism; we believe it has simplified other necessary proofs too and will perform further evaluation.
Although we cannot express mutually-recursive bindings in the core syntax here, we can express them as a notation on the surface syntax at the expense of potentially duplicating expressions.

\section{Extraction}
\label{s:extraction}

Pipit can generate executable code which is suitable for real-time execution on embedded devices.
The code extraction is implemented in a currently-unverified transform that takes a deeply embedded representation of a Pipit expression and generates an imperative representation of the program.
\fstar{} can generate C code from a subset of the language called \lowstar{}~\cite{protzenko2017verified}.
The result of our translation to imperative code fits in this subset.
During code extraction, we use \fstar{}'s tactic support~\cite{martinez2019meta} to fully normalise the translation to imperative code, similar to staged compilation.

% For the simple examples in \autoref{ss:tut:bibo}, we probably could have proved the properties directly on the generated code.
% In the future, however, we intend to support the abstraction of streams through \emph{contracts}, which require a relational semantics rather than a functional one.

% noextract
% let expr = Sugar.run2 Pump.controller'

% noextract
% let system: Pipit.Exec.Exp.xexec expr =
%   XX.exec_of_exp expr

% [@@(Tac.postprocess_with XL.tac_extract)]
% let reset = XL.mk_reset system

% [@@(Tac.postprocess_with XL.tac_extract)]
% let step (inp: input) = XL.mk_step system (inp.estop, (inp.level_low, ()))
% This allows us to generate C code which executes in bounded space on embedded devices.

\section{Evaluation}
\label{s:coffee-machine}

To demonstrate the feasibility of Pipit, we have implemented and verified a simple controller.
This system controls a water-flow solenoid to fill the reservoir of a coffee machine with multiple safeguards to reduce the risk of flooding.
The controller has two boolean inputs: the \emph{stop} switch and the \emph{low level} indicator; it returns a boolean indicating whether to engage the solenoid.
The stop switch indicates whether the reservoir's lid is open or closed; the system should never operate while the lid is open as water could spill out.
The second safeguard is that the solenoid should not be engaged for more than a minute or so as this may indicate a leak; if so, the controller enters a terminal \emph{stuck} state.
Finally, to avoid banging the solenoid too often, the controller waits for ten seconds of low water level before trying to engage.

\newcommand\estop{\textit{stop}}
\newcommand\low{\textit{low}}
\newcommand\soltry{\textit{try}}
\newcommand\stuck{\textit{stuck}}
\newcommand\solen{\textit{engage}}
\begin{tabbing}
  @MM@\= @let@ engage \= @MMMM@ \= \kill
  @let@ $\mbox{reservoir}$ (\estop{} \low{}: stream $\BB$): stream $\BB$ = \\
  \> @let@ \soltry{} \> = true\_for \textit{TEN\_SECONDS} (not \estop{} $\wedge$ \low{}) @in@ \\
  \> @let@ \stuck{} \> = once (true\_for \textit{ONE\_MINUTE} \soltry{}) @in@ \\
  \> @let@ \solen{} \> = \soltry{} $\wedge$ not \stuck{} @in@ \\
  \> @check@ (\solen{} $\implies$ not \estop{}); \\
  \> @check@ (\solen{} $\implies$ \low{}); \\
  \> \solen{}
\end{tabbing}

As with the examples in \autoref{ss:tut:bibo}, the above properties can be inductively verified by \fstar{}'s proof automation.
We can also generate executable C code, which runs in real-time on a microcontroller in the first author's kitchen.

@@

Earlier work verifying reactive systems used model-checking via explicit state-space exploration~\cite{raymond2008synchronous}; more recent work uses SMT solvers to find inductive proofs~\cite{hagen2008scaling,champion2016kind2} or check refinement types~\cite{chen2022synchronous}.
These model-checkers have definite advantages over the general-purpose-prover approach offered here: they can often generate concrete counterexamples that disprove properties; they can also implement advanced invariant generation techniques such as ICE~\cite{garg2014ice} and IC3/PDR~\cite{bradley2011sat,een2011efficient}.
However, these special-purpose model-checkers do not provide much assurance that the semantics they use for proofs matches the compiled code.
We believe that once Pipit's imperative code generation is verified, Pipit will have a stronger assurance case.

Early work embedding a denotational semantics of Lucid Synchrone in an interactive theorem prover focussed on the semantics itself, rather than proving programs~\cite{boulme2001clocked}.
There is ongoing work to construct a denotational semantics of Vélus for program verification~\cite{bourke2022towards}.
We believe that the hybrid SMT approach of \fstar{} will allow for a better mixture of automated proofs with manual proofs;
however, the trusted computing base of Pipit is much larger than Vélus, as we depend on all of \fstar{}, \lowstar{}'s C code extraction, the SMT solver, and our currently-unverified imperative code generator.

% Refinement types have also been used to verify reactive systems~\cite{chen2022synchronous}, but this work does not address the issue of correct compilation.

% cite also:
% hardware language in Agda \cite{harrison2021mechanized}
% Refinement types for Zélus \cite{chen2022synchronous}
% Invariant generation: ICE \cite{garg2014ice}; template-based \cite{kahsai2011instantiation}; PDR explanation \cite{een2011efficient}; IC3 first introduced \cite{bradley2011sat}

% Other compilers for Lustre-style languages, such as Vélus \cite{bourke2017formally} and Heptagon \cite{gerard2012modular} ensure that the generated code can be executed in bounded memory and that each step takes a bounded time.
% With a carefully chosen set of primitives, the core streaming operations shown here would ensure bounded memory and time.
% For practicality, however, we currently allow embedding arbitrary total functions from the \fstar{} meta-language, which are not necessarily bounded in time.
% (Is this the same as Lucid Synchrone and Zélus? \CITE)

\section{Conclusion}

We believe that these preliminary results show promise.
In the intermediate future, we intend to verify the imperative code generation.

TODO
Future work: verify imperative code generation; verify CSE; case studies: antilock braking; clocks \cite{caspi1995functional}; contracts \cite{champion2016kind2}.

% \pagebreak

### 2023-05-11 2023pipit/figures/core-grammar.tex
almost

\begin{figure}
  \[
  \begin{array}{lrlll}
    e, e' & := & v ~|~ x ~|~ e~e' \\
          & | & \xfby{v}{e} & | & \xthen{e}{e'} \\
          & | & \xrec{x}{e[x]} & | & \xlet{x}{e}{e'[x]} \\
          & | & \xcheck{e} \\
    \\
    v & := & \multicolumn{3}{l}{n \in \mathbb{N} ~|~ b \in \mathbb{B} ~|~ r \in \mathbb{R} ~|~ \hdots ~|~ \mlamX}
    \\
    \sigma & := & \sgl{\ov{x \mapsto v}} \\
    \Sigma & := & \sigma ~|~ \Sigma;\sigma \\
    %   \\
    % \tau, \tau' & := & \multicolumn{3}{l}{\mathbb{N} ~|~ \mathbb{B} ~|~ \hdots ~|~ \tau \to \tau'} \\
    % \Gamma & := & \cdot ~|~ x : \tau, \Gamma \\
    \end{array}
  \]
  \caption{Pipit language grammar, including expressions ($e$); values and host functions embedded from the \fstar{} meta-language ($v$); stores ($\sigma$); and non-empty stream histories ($\Sigma$).}
  \label{f:core-grammar}
\end{figure}

>>>>

\begin{figure}
  \[
  \begin{array}{lrlll}
    e, e' & := & v ~|~ x ~|~ e~e' \\
          & | & \xfby{v}{e} & | & \xthen{e}{e'} \\
          & | & \xrec{x}{e[x]} & | & \xcheck{e} \\
          & | & \multicolumn{3}{l}{\xlet{x}{e}{e'[x]}} \\
    \\
    v & := & \multicolumn{3}{l}{n \in \mathbb{N} ~|~ b \in \mathbb{B} ~|~ r \in \mathbb{R} ~|~ \hdots ~|~ \mlamX} \\
    % \sigma & := & \sgl{\ov{x \mapsto v}} \\
    % \Sigma & := & \sigma ~|~ \Sigma;\sigma \\
    %   \\
    % \tau, \tau' & := & \multicolumn{3}{l}{\mathbb{N} ~|~ \mathbb{B} ~|~ \hdots ~|~ \tau \to \tau'} \\
    % \Gamma & := & \cdot ~|~ x : \tau, \Gamma \\
    \end{array}
  \]
  \caption{Pipit language grammar including expressions~($e$); and values and host functions embedded from the \fstar{} meta-language~($v$).}
  \label{f:core-grammar}
\end{figure}

### 2023-05-11 2023pipit/shelve.tex
almost

>>>>

%%% META-THEORY

\begin{lemma}[no dependency, no difference]
  If $e$ does not have a direct depend on $x$, then any evaluation of $e$ cannot depend on the current value of $x$.

  For all expressions $e$ and well-scoped variables $x$,
  if $\neg \mbox{directly} ~e~ x$ and
  $\bigstep{\Sigma; (\sigma, x \mapsto v_x)}{e}{v}$
  then for any $v_x'$, evaluation is unchanged
  $\bigstep{\Sigma; (\sigma, x \mapsto v_x')}{e}{v}$
  \end{lemma}

  We also need:
  \begin{lemma}[Bigstep recursive XMu]
    If a recursive expression $\xrec{x}{e}$ evaluates to some sequence of values $v$, then evaluating the recursive expression doesn't depend on the last value in the sequence:

    if $\bigsteps{\Sigma}{\xrec{x}{e}}{v_0 v_1 \cdots v_{n-1} v_n }$
    then for any $v'$,
    $\bigsteps{\Sigma \triangleright^{\uparrow} (v_0 v_1 \cdots v_{n-1} v')}{e}{v_0 v_1 \cdots v_{n-1} v_n}$
  \end{lemma}

  % let bigstep_recursive_XMu (#outer #inner: nat) (e: exp)
  %   (streams: C.table (outer + 1) inner)
  %   (vs: C.vector value outer) (v v': value)
  %   (hBSmu: bigstep streams (XMu e) (v :: vs)):
  %     bigstep (C.table_map_append (C.table_of_values (v' :: vs)) streams) e (v :: vs) =

  \begin{lemma}[progress]
    if expression is causal, then for any valid stream history it should evaluate to a value:

    if $\typing{\Gamma}{e}{\tau}$ and $e \mbox{causal}$ and $\Gamma \vdash* \Sigma$, then exists $v$ such that
    $\bigstep{\Sigma}{e}{v}$ and $\mtyping{}{v}{\tau}$
  \end{lemma}

  \begin{lemma}[monotonic progress]
    if an expression $e$ evaluates to a sequence $vs$ under a stream history $\Sigma$ and a new stream store $\sigma$ is valid, then
    there exists some $v'$ such that
    $\bigsteps{\Sigma; \sigma}{e}{vs; v}$
  \end{lemma}

  \subsection{Meta-theory}
  \label{ss:meta-theory}

  We use a typed representation, which ensures an intrinsic variant of preservation.
  For programs that satisfy the causality restriction, we have proved a variant of the progress theorem.

  \begin{code}
  let bigstep_deterministic
    (#streams: list (C.row 'c))
    (#e: exp 'c 'a)
    (#v1 #v2: 'a)
    (hBS1: bigstep streams e v1) (hBS2: bigstep streams e v2):
      Lemma (ensures (v1 == v2)) (decreases hBS1) =
    bigstep_proof_equivalence hBS1 hBS2
  \end{code}

  Monotonicity

  Substitution-closure
  \begin{code}
  let rec bigstep_substitute_as_var (e p: exp)
    (streams1: C.table outer inner1)
    (streams2: C.table outer inner2)
    (vsp vsep: C.vector value outer)
    (hBSp: bigstep (C.table_map_append streams1 streams2) p vsp)
    (hBSep: bigstep (C.table_map_append streams1 streams2) (subst e inner1 p) vsep):
      Tot (bigstep (C.table_map_append streams1 (C.table_map_append (C.table_of_values vsp) streams2)) e vsep) (decreases hBSep) =

  \end{code}

  \section{Proving it with transition systems}

  Describe the translation from Pipit expressions to labelled transition systems.

  K-induction

  Statement of LTS soundness:
  \begin{code}
  let system_eval_complete'
      (#outer: nat) (#vars: nat)
      (e: exp { causal e /\ wf e vars })
      (streams: C.table outer vars)
      (vs: C.vector value outer)
      (hBS: bigstep streams e vs):
        Lemma (exists (s': option (state_of_exp e)).
          system_stepn'
            (system_of_exp e vars)
            (C.Table?._0 streams) vs s') =
      match outer with
      | 0 -> assert
        (system_stepn'
          (system_of_exp e vars)
          (C.Table?._0 streams) vs None)
      | _ -> system_eval_complete e streams vs hBS
  \end{code}
